Magnetised Polish doughnuts revisited

### 2017-03-21 paper.tex
last version

\author{Sergio Gimeno-Soler\inst{1}   \and Jos\'e A.~Font\inst{1,2} }
\institute{Departamento de Astronom\'{\i}a y Astrof\'{\i}sica, Universitat de Val\`encia, Dr. Moliner 50, 46100, Burjassot (Val\`encia), Spain.\\\email{sergiso@alumni.uv.es}
\and
Observatori Astron\`omic, Universitat de Val\`encia, C/ Catedr\'atico Jos\'e Beltr\'an 2, 46980, Paterna (Val\`encia), Spain. \\
\email{j.antonio.font@uv.es}
}

@@

\cite{Jaroszynski:1980} argued that the slope of the specific angular momentum should range between two limiting cases, namely $l = \mathrm{const.}$ and $\Omega = \mathrm{const}$. Following \citet{Qian:2009} we assume an angular momentum distribution ansatz given by  
\begin{equation}
l (r,\theta) = \left\{ \label{eq:ansatz} 
  \begin{array}{lr}
    l_0 \left(\frac{l_{\mathrm{K}}(r)}{l_0}\right)^{\beta}\sin^{2\gamma}{\theta} &  \text{for } r \geq r_{\mathrm{ms}}\\
    l_{\mathrm{ms}}(r)\sin^{2\gamma}{\theta} & \text{for } r < r_{\mathrm{ms}}
  \end{array}
\right.
\end{equation}
where constants $l_0$ and $l_{\mathrm{ms}}(r)$ are defined by $l_0 \equiv \eta l_{\mathrm{K}}(r_{\mathrm{ms}})$ and $l_{\mathrm{ms}}(r) \equiv l_0 [l_{\mathrm{K}}(r_{\mathrm{ms}})/l_0]^{\beta}$. Therefore, the model for the distribution of angular momentum has three free parameters, $\beta$ , $\gamma$ and $\eta$, whose range of variation is given by~\citep{Qian:2009}
\begin{equation}
0 \leq \beta \leq 1, \quad -1 \leq \gamma \leq 1, \quad 1 \leq \eta \leq \eta_{\mathrm{max}},
\end{equation}
with $\eta_{\mathrm{max}} = l_{\mathrm{K}}(r_{\mathrm{mb}})/l_{\mathrm{K}}(r_{\mathrm{ms}})$. In this paper, and as it is done for hydrodynamical discs in~\citet{Qian:2009}, we choose $\eta = \eta_{\mathrm{max}}$, and then we can write $l_0$ as $l_0 = l_{\mathrm{K}}(r_{\mathrm{mb}})$. For this choice of $\eta$, we can find the location of the cusp of the disc  within the range $r_{\mathrm{mb}} \leq r_{\mathrm{cusp}} \leq r_{\mathrm{ms}}$ (for $0 \leq \beta \leq 1$). Moreover, this value of $\eta$ guarantees that constant angular momentum discs ($\beta = \gamma = 0$) with their inner edge located at the cusp ($r_{\mathrm{in}} = r_{\mathrm{cusp}}$) have no outer boundary, i.e~they are infinite discs (see also~\cite{Font:2002}). \tf{Check that the last sentence is correct.}

@@

To construct our models of magnetised discs we follow the approach described in \citet{Qian:2009}. First, we find the radial location of the cusp and of the centre of the disc in the equatorial plane, $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$, defined as the solutions to the equation $l(r) - l_{\mathrm{K}} = 0$.
Next, we compute the partial derivatives of the potential, Eq.~\eqref{eq:potential}
\begin{equation}\label{eq:radial_der_pot}
\partial_r W = \partial_r \ln|u_t| - \frac{\Omega \partial_rl}{1 - \Omega l}\,,
\end{equation}
and
\begin{equation}\label{eq:polar_der_pot}
\partial_{\theta} W = \partial_{\theta} \ln|u_t| - \frac{\Omega \partial_{\theta}l}{1 - \Omega l}\,.
\end{equation}
Then, we integrate the radial partial derivative of the potential along the segment $[r_{\mathrm{cusp}}, r_{\mathrm{c}}]$ (assuming $W_{\mathrm{cusp}} = 0$) at the equatorial plane, thus obtaining the equatorial distribution of the potential between $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$
\begin{equation}\label{eq:equatorial_pot}
W_{\mathrm{eq}}(r) = \int^{r_{\mathrm{c}}}_{r_{\mathrm{cusp}}}\left(\partial_r \ln|u_t| - \frac{\Omega \partial_rl}{1 - \Omega l}\right).
\end{equation}
Following~\citet{Qian:2009} (see also~\citet{Jaroszynski:1980}) we can divide equations \eqref{eq:radial_der_pot} and \eqref{eq:polar_der_pot} to obtain
\begin{equation}\label{eq:F}
F(r, \theta) = -\frac{\partial_r W}{\partial_{\theta} W} = \frac{\mathrm{d}\theta}{\mathrm{d}r}\,,
\end{equation}
where function $F(r, \theta)$ is known in closed form for the Kerr metric once an assumption for the angular momentum distribution has been made (cf.~Eq.~(\ref{eq:ansatz})). For this reason, Eq.~(\ref{eq:F}) also takes the form of an ordinary differential equation for the surfaces of constant pressure, $\theta=\theta(r)$, which, upon integration, yields the location of those surfaces.

@@

Next, we choose all the initial radial values for the integration of Eq.~\eqref{eq:F} to lie between $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$ ($\theta = \pi / 2$). Since we are only interested in the equipressure surfaces inside the Roche lobe of the disc, our choice of initial values provides us a mapping of the equipotential surfaces of the torus. Given that we have already obtained both the equipotential surfaces which cross the segment $[r_{\mathrm{cusp}}, r_{\mathrm{c}}]$ and the potential \tf{pressure?} distribution there, we can obtain the potential distribution for the torus.
Once we have the potential distribution, we can find the gas pressure at the centre of the disc from Eq.~\eqref{eq:enthalpy_eq} \tf{we need to discuss this paragraph to see if we can improve the writing}
\begin{equation}
p_{\mathrm{c}} = w_{\mathrm{c}}(W_{\mathrm{in}} - W_{\mathrm{c}})\left(\frac{\kappa}{\kappa - 1} + \frac{\lambda
}{\lambda
 - 1}\frac{1}{\beta_{\mathrm{m}_{\mathrm{c}}}}\right)^{-1},
\end{equation}
where $w_{\mathrm{c}}$ is the enthalpy at the centre and
\begin{equation}
\label{eq:beta_eq}
\beta_{\mathrm{m}} = p/p_{\mathrm{m}},
\end{equation}
is the magnetisation parameter ($\beta_{\mathrm{m}_{\mathrm{c}}}$ being the magnetisation parameter at the centre of the disc). Using this definition, we can find the magnetic pressure at the centre,
\begin{equation}
p_{\mathrm{m_{\mathrm{c}}}} = p_{\mathrm{c}}/\beta_{\mathrm{m}_{\mathrm{c}}}\,.
\end{equation}
With both pressures known at the centre, fluid and magnetic, we can now find the constants $K$ and $K_{\mathrm{m}}$ using equations \eqref{eq:eos_fluid} and \eqref{eq:eos_mag}. Therefore, for a given inner radius of the disc $r_{\mathrm{in}}$ we can obtain the potential $W_{\mathrm{in}}$. We now finally have all the ingredients required to find the enthalpy distribution (Eq.~\eqref{eq:enthalpy_eq}), the fluid pressure and magnetic pressure distributions (Eqs.~\eqref{eq:eos_fluid} and \eqref{eq:eos_mag}), and the density distribution (Eq.~\eqref{eq:density_eq}). \tf{which equation?}

@@

Next, we show in Fig.~\ref{more-profile} the corresponding radial profiles at the equatorial plane for the models with black hole spin $a=0.9$. We consider the case $\beta_{\mathrm{m}_{\mathrm{c}}}=10^{-3}$ (highest magnetisation) and the same combination of the $\gamma$ and $\beta$ parameters that we employed in Fig.~\ref{models}. Therefore, this plot depicts the radial profiles of the models occupying the central column of Fig.~\ref{models}. These profiles allow for a clearer quantification of the radial extent of the discs with $\gamma$ and $\beta$. As mentioned in the description of Fig.~\ref{models} as $\gamma$ and $\beta$ increase from 0 to 0.9 (compare black, blue, and green curves) the discs become gradually smaller. At the same time, the radial location of the density maximum increases \tf{decreases?} and the maximum value of the central density (and pressure) decreases \tf{increases?}. It is worth noticing that the radial profiles of some of the models overlap below a certain radius. More precisely, the black and red curves overlap below $r\sim r_{\mathrm{c}}$ (which corresponds to $\rho \sim 1$) and also the green and orange curves. In addition, this figure clearly shows that, besides modifying the thickness of the disc, $\beta$ is the parameter which determines the value of the density (and pressure) maximum. On the other hand, the parameter $\gamma$ is only relevant for controlling the radial size of the discs.

@@

The dependence of the location of the maximum of the magnetic pressure with $\beta_{\mathrm{m}_{\mathrm{c}}}$ is shown in the right panel of Fig.~\ref{max-vs-magnetisation}. While this location also decreases with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$, the constant value achieved for values above $\beta_{\mathrm{m}_{\mathrm{c}}}\sim 10^{3}$ does depend on the specific values of $\beta$ and $\gamma$, contrary to what happens with the fluid pressure. It must be noted that the location of the maximum of the magnetic pressure is nearly identical for all models considered when $\beta_{\mathrm{m}_{\mathrm{c}}} \sim $ \tf{indicate approximate value}. At this value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ all models seem to cross, as it is more clearly shown in the inset of the right panel of Fig.~\ref{max-vs-magnetisation}.

As a final remark we note that in the left and middle panels of Fig.~\ref{max-vs-magnetisation} we do not show the data for models with $\beta \neq \gamma$ as they coincide with the respective models with the same value of $\beta$. Moreover, the middle and right panels only show the data for $a=0.99$ because changing the value of the spin parameter of the black hole only yields a change of scale. Therefore, if we choose the range of the graph accordingly, the plots are identical irrespective of the value of $a$. 
\tf{rephrase last sentence better?}

>>>>

\author{Sergio Gimeno-Soler\inst{1}   \and Jos\'e A.~Font\inst{1,2} }
\institute{Departamento de Astronom\'{\i}a y Astrof\'{\i}sica, Universitat de Val\`encia, Dr. Moliner 50, 46100, Burjassot (Val\`encia), Spain.\\\email{sergio.gimeno@uv.es}
\and
Observatori Astron\`omic, Universitat de Val\`encia, C/ Catedr\'atico Jos\'e Beltr\'an 2, 46980, Paterna (Val\`encia), Spain. \\
\email{j.antonio.font@uv.es}
}

@@

\cite{Jaroszynski:1980} argued that the slope of the specific angular momentum should range between two limiting cases, namely $l = \mathrm{const.}$ and $\Omega = \mathrm{const}$. Following \citet{Qian:2009} we assume an angular momentum distribution ansatz given by  
\begin{equation}
l (r,\theta) = \left\{ \label{eq:ansatz} 
  \begin{array}{lr}
    l_0 \left(\frac{l_{\mathrm{K}}(r)}{l_0}\right)^{\beta}\sin^{2\gamma}{\theta} &  \text{for } r \geq r_{\mathrm{ms}}\\
    l_{\mathrm{ms}}(r)\sin^{2\gamma}{\theta} & \text{for } r < r_{\mathrm{ms}}
  \end{array}
\right.
\end{equation}
where constants $l_0$ and $l_{\mathrm{ms}}(r)$ are defined by $l_0 \equiv \eta l_{\mathrm{K}}(r_{\mathrm{ms}})$ and $l_{\mathrm{ms}}(r) \equiv l_0 [l_{\mathrm{K}}(r_{\mathrm{ms}})/l_0]^{\beta}$. Therefore, the model for the distribution of angular momentum has three free parameters, $\beta$ , $\gamma$ and $\eta$, whose range of variation is given by~\citep{Qian:2009}
\begin{equation}
0 \leq \beta \leq 1, \quad -1 \leq \gamma \leq 1, \quad 1 \leq \eta \leq \eta_{\mathrm{max}},
\end{equation}
with $\eta_{\mathrm{max}} = l_{\mathrm{K}}(r_{\mathrm{mb}})/l_{\mathrm{K}}(r_{\mathrm{ms}})$. In this paper, and as it is done for hydrodynamical discs in~\citet{Qian:2009}, we choose $\eta = \eta_{\mathrm{max}}$, and then we can write $l_0$ as $l_0 = l_{\mathrm{K}}(r_{\mathrm{mb}})$. For this choice of $\eta$, we can find the location of the cusp of the disc  within the range $r_{\mathrm{mb}} \leq r_{\mathrm{cusp}} \leq r_{\mathrm{ms}}$ (for $0 \leq \beta \leq 1$). Moreover, this value of $\eta$ guarantees that constant angular momentum discs ($\beta = \gamma = 0$) with their inner edge located at the cusp ($r_{\mathrm{in}} = r_{\mathrm{cusp}}$) have no outer boundary, i.e~they are infinite discs (see also~\cite{Font:2002}).

@@

To construct our models of magnetised discs we follow the approach described in \citet{Qian:2009}. First, we find the radial location of the cusp and of the centre of the disc in the equatorial plane, $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$, defined as the solutions to the equation $l(r) - l_{\mathrm{K}} = 0$.
Next, we compute the partial derivatives of the potential, Eq.~\eqref{eq:potential}
\begin{equation}\label{eq:radial_der_pot}
\partial_r W = \partial_r \ln|u_t| - \frac{\Omega \partial_rl}{1 - \Omega l}\,,
\end{equation}
and
\begin{equation}\label{eq:polar_der_pot}
\partial_{\theta} W = \partial_{\theta} \ln|u_t| - \frac{\Omega \partial_{\theta}l}{1 - \Omega l}\,.
\end{equation}
Then, we integrate the radial partial derivative of the potential along the segment $[r_{\mathrm{cusp}}, r_{\mathrm{c}}]$ (assuming $W_{\mathrm{cusp}} = 0$) at the equatorial plane, thus obtaining the equatorial distribution of the potential between $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$
\begin{equation}\label{eq:equatorial_pot}
W_{\mathrm{eq}}(r) = \int^{r_{\mathrm{c}}}_{r_{\mathrm{cusp}}}\left(\partial_r \ln|u_t| - \frac{\Omega \partial_rl}{1 - \Omega l}\right).
\end{equation}
Following~\citet{Qian:2009} (see also~\citet{Jaroszynski:1980}) we can divide equations \eqref{eq:radial_der_pot} and \eqref{eq:polar_der_pot} to obtain
\begin{equation}\label{eq:F}
F(r, \theta) = -\frac{\partial_r W}{\partial_{\theta} W} = \frac{\mathrm{d}\theta}{\mathrm{d}r}\,,
\end{equation}
where function $F(r, \theta)$ is known in closed form for the Kerr metric once an assumption for the angular momentum distribution has been made (cf.~Eq.~(\ref{eq:ansatz})). For this reason, Eq.~(\ref{eq:F}) also takes the form of an ordinary differential equation for the surfaces of constant potential, $\theta=\theta(r)$, which, upon integration, yields the location of those surfaces. Note that in~\citet{Qian:2009} these are surfaces of constant {\it pressure} instead, since their discs are purely hydrodynamical.

@@

Next, we choose all the initial radial values for the integration of Eq.~\eqref{eq:F} to lie between $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$ ($\theta = \pi / 2$). Since we are only interested in the equipotential surfaces inside the Roche lobe of the disc, our choice of initial values provides us a mapping of the equipotential surfaces of the torus. Given that we have already obtained both the equipotential surfaces $\theta(r)$ which cross the segment $[r_{\mathrm{cusp}}, r_{\mathrm{c}}]$ at the equatorial plane and the values of the potential in that segment, we can obtain the complete potential distribution for the torus (outside of that segment). Once we have the potential distribution, we can find the gas pressure at the centre of the disc from Eq.~\eqref{eq:enthalpy_eq},
\begin{equation}
p_{\mathrm{c}} = w_{\mathrm{c}}(W_{\mathrm{in}} - W_{\mathrm{c}})\left(\frac{\kappa}{\kappa - 1} + \frac{\lambda
}{\lambda
 - 1}\frac{1}{\beta_{\mathrm{m}_{\mathrm{c}}}}\right)^{-1},
\end{equation}
where $w_{\mathrm{c}}$ is the enthalpy at the centre and
\begin{equation}
\label{eq:beta_eq}
\beta_{\mathrm{m}} = p/p_{\mathrm{m}},
\end{equation}
is the magnetisation parameter ($\beta_{\mathrm{m}_{\mathrm{c}}}$ being the magnetisation parameter at the centre of the disc). Using this definition, we can find the magnetic pressure at the centre,
\begin{equation}
p_{\mathrm{m_{\mathrm{c}}}} = p_{\mathrm{c}}/\beta_{\mathrm{m}_{\mathrm{c}}}\,.
\end{equation}
With both pressures known at the centre, fluid and magnetic, we can now find the constants $K$ and $K_{\mathrm{m}}$ using equations \eqref{eq:eos_fluid} and \eqref{eq:eos_mag}. Therefore, for a given inner radius of the disc $r_{\mathrm{in}}$ we can obtain the potential $W_{\mathrm{in}}$. We now finally have all the ingredients required to find the enthalpy distribution (Eq.~\eqref{eq:enthalpy_eq}), the fluid pressure and magnetic pressure distributions (Eqs.~\eqref{eq:eos_fluid} and \eqref{eq:eos_mag}), and the density distribution, which can be trivially obtained inverting the barotropic relation, $p = K \rho^{\kappa}$.

@@

Next, we show in Fig.~\ref{more-profile} the corresponding radial profiles at the equatorial plane for the models with black hole spin $a=0.9$. We consider the case $\beta_{\mathrm{m}_{\mathrm{c}}}=10^{-3}$ (highest magnetisation) and the same combination of the $\gamma$ and $\beta$ parameters that we employed in Fig.~\ref{models}. Therefore, this plot depicts the radial profiles of the models occupying the central column of Fig.~\ref{models}. These profiles allow for a clearer quantification of the radial extent of the discs with $\gamma$ and $\beta$. As mentioned in the description of Fig.~\ref{models} as $\gamma$ and $\beta$ increase from 0 to 0.9 (compare black, blue, and green curves) the discs become gradually smaller. At the same time, the radial location of the density maximum increases and the maximum value of the central density (and pressure) decreases. It is worth noticing that the radial profiles of some of the models overlap below a certain radius. More precisely, the black and red curves overlap below $r\sim r_{\mathrm{c}}$ (which corresponds to $\rho \sim 1$) and also the green and orange curves. In addition, this figure clearly shows that, besides modifying the thickness of the disc, $\beta$ is the parameter which determines the value of the density (and pressure) maximum. On the other hand, the parameter $\gamma$ is only relevant for controlling the radial size of the discs.

@@

The dependence of the location of the maximum of the magnetic pressure with $\beta_{\mathrm{m}_{\mathrm{c}}}$ is shown in the right panel of Fig.~\ref{max-vs-magnetisation}. While this location also decreases with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$, the constant value achieved for values above $\beta_{\mathrm{m}_{\mathrm{c}}}\sim 10^{3}$ does depend on the specific values of $\beta$ and $\gamma$, contrary to what happens with the fluid pressure. It must be noted that the location of the maximum of the magnetic pressure is nearly identical for all models considered when $\log_{10}\beta_{\mathrm{m}_{\mathrm{c}}} \sim 0.5$. At this value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ all models seem to cross, as it is more clearly shown in the inset of the right panel of Fig.~\ref{max-vs-magnetisation}.

As a final remark we note that in the left and middle panels of Fig.~\ref{max-vs-magnetisation} we do not show the data for models with $\beta \neq \gamma$ as they coincide with the respective models with the same value of $\beta$. Moreover, the middle and right panels only show the data for $a=0.99$ because changing the value of the spin parameter of the black hole only yields a change of scale. Therefore, if we chose the range of the graph accordingly, the plots would be identical (same relative differences between the curves) irrespective of the value of $a$.
